### 2018-02-14 submission/manuscript.Rmd
update results to bring in line with upated fig legend

***Lower Bacterial Diversity is Associated with Increased OR of Tumors:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in OR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the OR for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and 16S variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased. For colon tissue, there were no significant changes in OTU richness, evenness, or Shannon diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C]. We next analyzed the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) colon tissue samples. For individuals at either an adenoma or carcinoma stage of disease there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5]. Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were identified in unmatched tissue samples, for those at either adenoma or carcinoma stage of CRC [Table S6 & S7]. For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7]. These tissue results suggest that the microbiota within an individual are similar to each other regardless of disease status.

***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in OR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Porphyromonas* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* and *Parvimonas* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with the gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased OR genera between both stages and sample site.

When observing ORs for adenoma between genera from stool or colon tissue with a P-value less than 0.05 there was almost no overlap and when they were similar the OR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the adenoma associated genera ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Pseudomonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*) had an increased OR for both colon tissue and stool samples [Table S10 & S11]. The genera with the highest increased OR for carcinoma in tissue was *Lepttrichia* while in stool it was *Peptostreptococcus* [Table S10 & S11].

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased the OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their OR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the full and select genera models, for both tissue and stool studies [Figure S2-S3]. When analyzing the tests sets that were comprised of genera data from other studies, both models had a similar ability to detect individuals with adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6]. Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models [Figure 4 & 5].

In stool, the most common genera in the top 10 most important variables, in the full community models using the first approach, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B]. Yet, none were present in the majority of studies and *Fusobacterium* was the only genus present in the adenoma stage of CRC [Figure 6A & B]. For the second approach that utilized full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].

Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S7]. The vast majority of the top 10 model variables for the genera- and OTU-based models using colon tissue tended to be study specific. Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure 6 & S7]. This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S7B-C & S7E-F]. This suggests that either the colon tissue microbiota is study and person dependent or that kit and/or other types of contamination associated with low biomass samples may be skewing the results.

***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls. When assessing the power of each study at different effect sizes the majority of studies achieved 80% power to detect a 30% or greater difference between groups [Figure 7A & B]. No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 7A & B].  In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 7C]

>>>>

***Lower Bacterial Diversity is Associated with Increased OR of Tumors:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively). We next tested whether the detectable differences in community significantly increased in OR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased OR for carcinoma  (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the OR for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 1]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1].

Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and 16S variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased. For colon tissue, there were no significant changes in OTU richness, evenness, or Shannon diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). We next analyzed the OR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) colon tissue samples. For individuals at either an adenoma or carcinoma stage of disease there was no significant change in OR based on lower than median values for OTU richness, evenness, and Shannon diversity [Figure S1 & Table S2]. Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were identified in unmatched tissue samples, for those at either adenoma or carcinoma stage of CRC [Table S1]. For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S1]. These tissue results suggest that the microbiota within an individual are similar to each other regardless of disease status.

***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in OR. For both tissue and stool samples only ORs associated with an increase or decrease in carcioma tumors were significant after multiple comparison correction [Table S3 & S4]. Mouth-associated genera significantly increased the ORs of carcinoma for stool samples and included *Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Table S3]. Conversely, mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4]. The significant OR taxa in tissue were *Dorea* (decreased OR) and *Weissella* (increased OR). Overall, there was little direct overlap in increased or decreased OR genera between both tumor type and sample site.

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased the OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on significance after multiple comparison correction. If no taxa were significant after multiple comparison correction then no model for that specific grouping (i.e. adenoma stool) was analyzed. We first tested three model AUCs. These models were created using Random Forest and where either all genera, all OTUs, or significant OR taxa only. Next, the all genera models and any significant OR taxa models were tested across all studies that were not used to train the model. For stool, all models used had similar AUCs [Figure 2]. Although for adenoma and unmatched carcinoma this trend held, there were large differences in matched tissue based on whether all genera or OTUs were utilized [Figure S2]. When analyzing the tests sets that were comprised of genera data from other studies, both the all genera and singificant OR taxa only models had a similar ability to detect individuals with carcinomas, for both stool and tissue samples [Figure 3-S3].

In stool, the most common genera in the top 10 most important variables, in the full community models using all genera-based models, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 4]. Regardless of sample type, mouth-associated genera were present in models for carcinomas [Figure 4B & D]. Yet, none were present in the majority of studies and none were present in the adenoma models [Figure 4A & B]. For the full community OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 4C & 4D].

Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S4]. The vast majority of the top 10 model variables for the full community genera- and OTU-based models using colon tissue tended to be study specific. Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure S4]. This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S4B-C & S4E-F]. This suggests that either the colon tissue microbiota is study or person dependent, that kit and/or other types of contamination associated with low biomass samples may be skewing the results, or that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations.

***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls. When assessing the power of each study at different effect sizes the majority of studies achieved 80% power to detect a 30% or greater difference between groups [Figure 5A & B]. No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 5A & B].  In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 5C]

### 2018-02-14 submission/manuscript.tex
update results to bring in line with upated fig legend

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased OR
of Tumors:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in OR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overall median in evenness resulted in a
significantly increased RR for carcinoma (OR = 1.66 (1.2 - 2.3), P-value
= 0.0021) and adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) while a
decrease versus the overall median in Shannon diversity only increased
the OR for carcinoma (OR = 1.61 (1.14 - 2.28), P-value = 0.0069)
{[}Figure 2{]}. Using the Bray-Curtis distance metric and PERMANOVA, it
was also possible to identify significant bacterial community changes,
in specific studies, for both carcinoma-associated and
adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

Using similar transformations for tissue samples, linear mixed-effect
models were used on the transformed combined data to control for study,
re-sampling of the same individual, and 16S variable region to test
whether OTU richness, evenness, or Shannon diversity changed in a
step-wise manner as disease severity increased. For colon tissue, there
were no significant changes in OTU richness, evenness, or Shannon
diversity as disease severity progressed from control to adenoma to
carcinoma (P-value \textgreater{} 0.05) {[}Figure 1B \& C{]}. We next
analyzed the RR, for matched (unaffected tissue and an adenoma or
carcinoma from the same individual) and unmatched (control and adenoma
or carcinoma tissue not from the same individual) colon tissue samples.
For individuals at either an adenoma or carcinoma stage of disease there
was no significant change in RR based on lower than median values for
OTU richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar
to stool samples, significant differences in bacterial community,
assessed by PERMANOVA, were identified in unmatched tissue samples, for
those at either adenoma or carcinoma stage of CRC {[}Table S6 \& S7{]}.
For studies with matched samples no differences in bacterial community
were observed when assessed with PERMANOVA {[}Table S6 \& S7{]}. These
tissue results suggest that the microbiota within an individual are
similar to each other regardless of disease status.

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
OR of Tumor:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered OR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in OR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
OR. Mouth-associated genera were commonly found in the top 5 genera
associated with an increased OR of having an adenoma
(\emph{Porphyromonas} {[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]})
and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} and \emph{Parvimonas} {[}Figure 3D{]}) for both
stool and colon tissue samples. Conversely, genera commonly associated
with the gastrointestinal tract were correlated with a decreased OR for
both adenoma and carcinoma for both stool and colon tissue samples
{[}Figure 3{]}. Even though mouth-associated genera were identified
across disease stage, there was little direct overlap of the top 5
increased or decreased OR genera between both stages and sample site.

When observing ORs for adenoma between genera from stool or colon tissue
with a P-value less than 0.05 there was almost no overlap and when they
were similar the OR was in opposite directions (e.g. \emph{Lactococcus})
{[}Table S8 \& S9{]}. Many of the adenoma associated genera ORs with a
P-value under 0.05 for colon tissue are also highly prevalent in
contamination, specifically, \emph{Novosphingobium}, \emph{Pseudomonas},
and \emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma
stage of disease, certain mouth-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}) had an increased OR for both colon tissue and stool
samples {[}Table S10 \& S11{]}. The genera with the highest increased OR
for carcinoma in tissue was \emph{Lepttrichia} while in stool it was
\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific genera increased the OR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on their OR and P-value significance and used two
approaches to test this question. The first approach used genus level
data and tested for differences in AUC between all genera and selected
genera. A single study was used for training the model prior to testing
on all other studies and this was repeated for every study in the
meta-analysis. The second approach used OTU level data and tested for a
generalized decrease in the 10-fold cross validation (CV) model AUC
which is a common approach used to guard against over-fitting. This was
applied across study and the AUC of the all OTUs model was compared
against the model that used only OTUs that taxonomically classified to
selected genera.

For the first approach using the genera-based models, the training set
median AUC for model classification was similar for both the full and
select genera models, for both tissue and stool studies {[}Figure
S2-S3{]}. When analyzing the tests sets that were comprised of genera
data from other studies, both models had a similar ability to detect
individuals with adenomas or carcinomas, with the select genera models
performing better in some instances {[}Figure S4-S6{]}. Conversely, the
second approach that used OTU-based models showed a slight decrease in
median AUC between the full and select models {[}Figure 4 \& 5{]}.

In stool, the most common genera in the top 10 most important variables,
in the full community models using the first approach, were
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
6A \& B{]}. Regardless of sample type, mouth-associated genera were
present in models for the carcinoma stage of CRC {[}Figure 6A \& B{]}.
Yet, none were present in the majority of studies and
\emph{Fusobacterium} was the only genus present in the adenoma stage of
CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
OTU-based models, \emph{Ruminococcaceae} was present in the top 10
consistently for both adenoma and carcinoma models while
\emph{Roseburia} was only present in many adenoma models and
\emph{Bacteroides} was present in the overwhelming majority of the
carcinoma models {[}Figure 6C \& 6D{]}.

Unlike the stool-based Random Forest models, the tissue-based models,
for the full genera from the first approach, showed no consistent
representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} in the top 10 most important
model variables across study {[}Figure S7{]}. The vast majority of the
top 10 model variables for the genera- and OTU-based models using colon
tissue tended to be study specific. Further, there was very little
overlap in the top 10 important variables between adenoma and carcinoma
stage models, regardless of whether colon tissue or stool was used
{[}Figure 6 \& S7{]}. This discordance between stool and colon tissue
samples also applies to the mouth-associated genera with one noticeable
skew being that \emph{Fusobacterium} and \emph{Fusobacteriaceae} occur
more often in the top 10 of matched versus unmatched colon tissue Random
Forest models {[}Figure S7B-C \& S7E-F{]}. This suggests that either the
colon tissue microbiota is study and person dependent or that kit and/or
other types of contamination associated with low biomass samples may be
skewing the results.

\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
Sizes:}} Next, we assessed how much confidence should be placed in the
reported outcomes from each individual study by calculating the ability
to detect a difference (power) and sample size needed for small, medium,
and large effect size differences between cases and controls. When
assessing the power of each study at different effect sizes the majority
of studies achieved 80\% power to detect a 30\% or greater difference
between groups {[}Figure 7A \& B{]}. No study that we analyzed had the
standard 80\% power to detect an effect size difference equal to or
below 10\% {[}Figure 7A \& B{]}. In order to achieve a power of 80\%,
for small effect sizes, studies used in our meta-analysis would need to
recruit over 1000 individuals for both the case and control arms
{[}Figure 7C{]}

>>>>

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased OR
of Tumors:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively). We next tested whether the
detectable differences in community significantly increased in OR of
having an adenoma or carcinoma. For fecal samples, a decrease versus the
overall median in evenness resulted in a significantly increased OR for
carcinoma (OR = 1.66 (1.2 - 2.3), P-value = 0.0021) and adenoma (OR =
1.3 (1.02 - 1.65), P-value = 0.035) while a decrease versus the overall
median in Shannon diversity only increased the OR for carcinoma (OR =
1.61 (1.14 - 2.28), P-value = 0.0069) {[}Figure 1{]}. Using the
Bray-Curtis distance metric and PERMANOVA, it was also possible to
identify significant bacterial community changes, in specific studies,
for both carcinoma-associated and adenoma-associated microbiota versus
control {[}Table S1{]}.

Using similar transformations for tissue samples, linear mixed-effect
models were used on the transformed combined data to control for study,
re-sampling of the same individual, and 16S variable region to test
whether OTU richness, evenness, or Shannon diversity changed in a
step-wise manner as disease severity increased. For colon tissue, there
were no significant changes in OTU richness, evenness, or Shannon
diversity as disease severity progressed from control to adenoma to
carcinoma (P-value \textgreater{} 0.05). We next analyzed the OR, for
matched (unaffected tissue and an adenoma or carcinoma from the same
individual) and unmatched (control and adenoma or carcinoma tissue not
from the same individual) colon tissue samples. For individuals at
either an adenoma or carcinoma stage of disease there was no significant
change in OR based on lower than median values for OTU richness,
evenness, and Shannon diversity {[}Figure S1 \& Table S2{]}. Similar to
stool samples, significant differences in bacterial community, assessed
by PERMANOVA, were identified in unmatched tissue samples, for those at
either adenoma or carcinoma stage of CRC {[}Table S1{]}. For studies
with matched samples no differences in bacterial community were observed
when assessed with PERMANOVA {[}Table S1{]}. These tissue results
suggest that the microbiota within an individual are similar to each
other regardless of disease status.

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
OR of Tumor:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered OR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in OR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
OR. For both tissue and stool samples only ORs associated with an
increase or decrease in carcioma tumors were significant after multiple
comparison correction {[}Table S3 \& S4{]}. Mouth-associated genera
significantly increased the ORs of carcinoma for stool samples and
included \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Porphyromonas},
and \emph{Peptostreptococcus} {[}Table S3{]}. Conversely,
mouth-associated genera were not significantly associated with an
increased OR for carcinoma in tissue samples {[}Table S4{]}. The
significant OR taxa in tissue were \emph{Dorea} (decreased OR) and
\emph{Weissella} (increased OR). Overall, there was little direct
overlap in increased or decreased OR genera between both tumor type and
sample site.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific genera increased the OR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on significance after multiple comparison correction.
If no taxa were significant after multiple comparison correction then no
model for that specific grouping (i.e.~adenoma stool) was analyzed. We
first tested three model AUCs. These models were created using Random
Forest and where either all genera, all OTUs, or significant OR taxa
only. Next, the all genera models and any significant OR taxa models
were tested across all studies that were not used to train the model.
For stool, all models used had similar AUCs {[}Figure 2{]}. Although for
adenoma and unmatched carcinoma this trend held, there were large
differences in matched tissue based on whether all genera or OTUs were
utilized {[}Figure S2{]}. When analyzing the tests sets that were
comprised of genera data from other studies, both the all genera and
singificant OR taxa only models had a similar ability to detect
individuals with carcinomas, for both stool and tissue samples {[}Figure
3-S3{]}.

In stool, the most common genera in the top 10 most important variables,
in the full community models using all genera-based models, were
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
4{]}. Regardless of sample type, mouth-associated genera were present in
models for carcinomas {[}Figure 4B \& D{]}. Yet, none were present in
the majority of studies and none were present in the adenoma models
{[}Figure 4A \& B{]}. For the full community OTU-based models,
\emph{Ruminococcaceae} was present in the top 10 consistently for both
adenoma and carcinoma models while \emph{Roseburia} was only present in
many adenoma models and \emph{Bacteroides} was present in the
overwhelming majority of the carcinoma models {[}Figure 4C \& 4D{]}.

Unlike the stool-based Random Forest models, the tissue-based models,
for the full genera from the first approach, showed no consistent
representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} in the top 10 most important
model variables across study {[}Figure S4{]}. The vast majority of the
top 10 model variables for the full community genera- and OTU-based
models using colon tissue tended to be study specific. Further, there
was very little overlap in the top 10 important variables between
adenoma and carcinoma stage models, regardless of whether colon tissue
or stool was used {[}Figure S4{]}. This discordance between stool and
colon tissue samples also applies to the mouth-associated genera with
one noticeable skew being that \emph{Fusobacterium} and
\emph{Fusobacteriaceae} occur more often in the top 10 of matched versus
unmatched colon tissue Random Forest models {[}Figure S4B-C \& S4E-F{]}.
This suggests that either the colon tissue microbiota is study or person
dependent, that kit and/or other types of contamination associated with
low biomass samples may be skewing the results, or that multiple
microbes could act as the inflammatory stimulus needed to exacerbate
mutations.

\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
Sizes:}} Next, we assessed how much confidence should be placed in the
reported outcomes from each individual study by calculating the ability
to detect a difference (power) and sample size needed for small, medium,
and large effect size differences between cases and controls. When
assessing the power of each study at different effect sizes the majority
of studies achieved 80\% power to detect a 30\% or greater difference
between groups {[}Figure 5A \& B{]}. No study that we analyzed had the
standard 80\% power to detect an effect size difference equal to or
below 10\% {[}Figure 5A \& B{]}. In order to achieve a power of 80\%,
for small effect sizes, studies used in our meta-analysis would need to
recruit over 1000 individuals for both the case and control arms
{[}Figure 5C{]}

### 2018-02-14 submission/supplemental.Rmd
update results to bring in line with upated fig legend

**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stooland Tissue**

@@

**Table S3: OR of Tumor in Stool by Individual Genera with Initial P-Value under 0.05**

@@

**Table S4: OR of Tumor in Tissue by Individual Genera with Initial P-Value under 0.05**

>>>>

**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool and Tissue**

@@

**Table S3: OR in Stool of Tumor by Individual Genera with Initial P-Value under 0.05**

@@

**Table S4: OR in Tissue of Tumor by Individual Genera with Initial P-Value under 0.05**
